\chapter{Background}
% PLAN: 4k words

This chapter describes CHERI (\cref{chap:bg:sec:cheri}) and RVV (\cref{chap:bg:sec:rvv}) to the detail required to understand the rest of the dissertation.
Both are described because this dissertation caters to multiple audiences: those who already understand CHERI, but are less familiar with vectors; and those already familiar with vector hardware but not with CHERI.
The final section describes two other vector models in a little more detail.
They are both proprietary, so we cannot learn from their hardware, but they are useful for contextualizing the software side of things.


\section{CHERI}\label{chap:bg:sec:cheri}
% Concept of a capability
In CHERI, addresses/pointers are replaced with capabilities: unforgeable tokens that provide \emph{specific kinds of access} to an \emph{address} within a \emph{range of memory}.
The above statement is enough to understand what capabilities contain\footnote{This is a slight simplification. For the purposes of vector memory accesses the \emph{otype} of a capability can be ignored, as any type other than \code{UNSEALED} cannot be dereferenced anyway.}:
\begin{itemize}
    \item Permission bits, to restrict access
    \item The \emph{cursor}, i.e. the address it currently points to
    \item The \emph{bounds}, i.e. the range of addresses this capability could point to
\end{itemize}
A great deal of work has gone into compressing capabilities down into a reasonable size (see \cite{woodruffCHERIConcentratePractical2019}, \todoref{diagram from TR-941?}), and using the magic of floating-point all of this data has been reduced to just 2x the architectural register size.
For example, on 64-bit RISC-V a standard capability is 128-bits long.
The rest of this dissertation assumes capabilities are 128-bits long for simplicity.

In order to keep track of what is and isn't a valid capability, registers and memory are both tagged.
Each 128-bit register and each aligned 128-bit region of memory has an associated tag bit, which denotes if its data encodes a valid capability\footnote{This has the side-effect that capabilities must be 128-bit aligned in memory.}.
If any non-capability data is written to any part of the region, the tag bit is zeroed out.
\todomark{Can you have a capability with tag=1, but the cursor is outside the bounds?}
As above, significant work has gone into the implementation to reduce the DRAM overhead of this method (see \cite{joannouEfficientTaggedMemory2017} for an example).


\todomark{Explain the three security properties from davisCheriABIEnforcingValid2019?}


\todomark{rewrite above section - just copied over some stuff previously from Emulation Investigation and it doesn't flow}


\subsection{CHERI-RISC-V ISA}
The Cambridge Computer Lab's TR-951 report\todocite{TR-951} describes the latest version of the CHERI architecture (CHERI ISAv8) and proposes applications to MIPS, x86-64, and RISC-V.
CHERI-RISC-V is a mostly straightforward set of additions to basic RISC-V ISAs.
It adds thirty-two general-purpose capability registers, thirty-two Special Capability Registers (SCRs), and many new instructions.

The new general-purpose capability registers are each of size \code{CLEN = 2 * XLEN} plus a tag bit.
These registers store compressed capabilities.
While there is always a logical distinction between the pre-existing \emph{integer} registers \code{x0-x31} and the \emph{capability} registers \code{cx0-cx31}, the architecture may store them in a Split or Merged register file.
A Split register file stores the integer registers separately from capability registers, so programs can manipulate them independently.
A Merged register file stores thirty-two registers of length \code{CLEN}, using the full width for the capability registers, and aliases the integer registers to the bottom \code{XLEN} bits.
Under a merged register file, writing to an integer register makes the capability counterpart invalid, so programs have to be more careful with register usage.

\todomark{diagram for split/merged register file?}

\todomark{All new CHERI instructions specify if their operands are integers or capabilities, and pull from the correct set of logical registers.}
\todomark{Non-CHERI enabled instructions default to integer representations, and in some cases can be switched to use capabilities using the Integer/Capability encoding mode flag}

Many of the new SCRs are intended to support the privileged ISA extensions for e.g. hypervisors or operating systems.
The emulator doesn't use these, so their SCRs are not listed here, but there are two highly relevant SCRs for all modes: the Program Counter Capability and the Default Data Capability.
The PCC supplements \todomark{replaces?} the program counter by adding more metadata, ensuring instruction fetches have the same security as normal loads and stores.


\subsection{Integer/Capability mode}
\todomark{Capability mode flag overrides behaviour of RISC-V }

\subsection{Capability Relocations}
\todomark{talk about ELF cap-relocs table, how they work and why}

%---------------------------------
%---------------------------------
%---------------------------------
\section{A brief history of vector processing}
Many vector implementations (Intel SSE/AVX, Arm's Advanced SIMD and Neon) use fixed-length vectors - e.g. 128-bit vectors which a program interprets as four 32-bit elements.
As the industry's desire for parallelism grew, new implementations had to be designed with longer vectors of more elements.
For example, Intel SSE/SSE2 (both 128-bit) was succeeded by AVX (128 and 256-bit), then AVX2 (entirely 256-bit), then AVX-512 (512-bit).
Programs built for one extension, and hence designed for a specific vector size, could not automatically take advantage of longer vectors.

Scalable vectors address this by not specifying the vector length, and instead calculating it on the fly.
Instead of hardcoding \enquote{this loop iteration uses a single vector of four 32-bit elements}, the program has to ask \enquote{how many 32-bit elements will this iteration use?}.
This gives hardware designers more freedom, letting them select a suitable hardware vector length for their power/timing targets, while guaranteeing consistent execution of programs on arbitrarily-sized vectors.
RVV uses a scalable vector model.

\include{1_20Background/rvv}

\section{Other vector models}
\subsection{Arm SVE}
\subsection{Intel SSE/AVX}